---
title: "Assignment 7: Bayesian General Linear Models"
author: "Your Name Here"
execute:
  echo: true
  warning: false
  message: false
  error: true
format: 
  html:
    df-print: kable
    self-contained: true
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
editor: source
---

**In this assignment, you will use the `penguins_bayes` data in the `bayesrules` package to build various models of penguin body mass (`body_mass_g`). We have a basic understanding that the average penguin weighs somewhere between 3500 and 4500 grams. A predictor of interest is penguin species: Adelie, Chinstrap, or Gentoo.**

```{r}
library(bayesrules)
penguin_data <- penguins_bayes
```

## Model 1

**Let's explore body mass (`body_mass_g`) as a function of flipper length (`flipper_length_mm`) and species (`species`).**

**a. Plot body mass against flipper length, colored by species.**

```{r}

```

**b. What pattern(s) do you observe, if any, in the above graph?**

Your response here

**c. Use `stan_glm()` to simulate the specified model.**

```{r}
#| results: "hide"


```

**d. What is the estimated model?**

$$\hat{y} = \beta_0 + \beta_1 \text{var1} + \beta_2 \text{var2} + ...$$ 

<!-- make sure you actually plug in the values for the betas -- use what you estimated in part c -->

**e. Construct visual diagnostics for your MCMC simulations.**

```{r}

```

**f. Comment on the diagnostics. Do you see any issues with our simulation?**

Your response here

**g. Construct the visual posterior predictive check for your model.**

```{r}

```

**h. How wrong is the model?**

Your response here.

**i. Plot the data with the posterior predictive 50% and 95% credible intervals.**

```{r}

```

**j. Find and interpret the scaled absolute median error.**

```{r}

```

**k. How accurate is the posterior predictive model?**

Your response here.

## Model 2

**Let's now explore body mass (`body_mass_g`) as a function of flipper length (`flipper_length_mm`), species (`species`), and the interaction between the two.**

**a. Use `stan_glm()` to simulate the specified model.**

```{r}
#| results: "hide"


```

**b. What is the estimated model?**

$$\hat{y} = \beta_0 + \beta_1 \text{var1} + \beta_2 \text{var2} + ...$$ 

<!-- make sure you actually plug in the values for the betas -- use what you estimated in part a -->

**c. Determine if this model is better than the previous model.**

Your response here.

<!-- this is totally a stretch question -- there's not a pre-defined right answer -->
<!-- you will want to research to figure out how to perform the comparison -->

**d. Do you recommend keeping the interaction in the model? Why or why not?**

Your response here.

<!-- this response should be based on your part c -->

